---
title: "BIOL3207_Assignment_2"
output: html_document
date: "2022-10-17"
---
Notes for assignments
```{r eval=F}
Notes:
  - Meta-analysis on ocen acidification effects on behaviour
  - On 5-6 reef fish -> Elevated CO2 + assess fish behaviour
    - Did not find strong evidence that increased acification had effects on behaviour
Findings:
  - Estimate overall effect of ocean acidification on behaviour and determine if these effects are general across studies conducting similar experiments;
  - Understand how variable the effect size is within the literature
  - What factors (biological, methodological, publication practices) explain variation in effect size
```

Link to repository: https://github.com/DeniseTsoi/3207_Assignment2

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **(1) Statistical Analysis and Interpretation (50%)**
### 1. Correct analysis of Clark et al. (2020) data (i.e., OA_activitydat_20190302_BIOL3207.csv) to generate the summary statistics (means, SD, N) for each of the fish species’ average activity for each treatment.
```{r}
library(readr)
library(tidyverse)
OA_data <- read_csv("./data/Assignment2/OA_activitydat_20190302_BIOL3207.csv")
clark_data <- read_csv("./data/Assignment2/clark_paper_data.csv")
OA_summary <- OA_data %>% group_by(species, treatment) %>%
              summarise(mean = mean(activity, na.rm = TRUE),
                        sd = sd(activity, na.rm = TRUE),
                        n = length(unique(animal_id))) %>%
              rename(Species = "species")
```

### 2. Through coding, merge the summary statistics generated from 1) with the metadata (i.e., clark_paper_data.csv) from Clark et al. (2020).
```{r}
OA_clark <- cbind(clark_data, OA_summary)
OA_clark1 <- pivot_wider(OA_clark, names_from = treatment,
                     names_glue = "{treatment}_{.value}",
                     values_from = c("mean", "sd", "n"))
```

### 3. Through coding, correctly merge the combined summary statistics and metadata from Clark et al. (2020) (output from 1 & 2) into the larger meta-analysis dataset (i.e., ocean_meta_data.csv).
```{r}
ocean_meta <- read_csv("./data/Assignment2/ocean_meta_data.csv")

## Renaming some columns so that they match with ocean_meta
OA_clark2 <- OA_clark1 %>% rename("oa.mean" = CO2_mean,
                            "oa.sd" = CO2_sd,
                            "oa.n" = CO2_n,
                            "ctrl.mean" = control_mean,
                            "ctrl.sd" = control_sd,
                            "ctrl.n" = control_n)

# Reorder col names based on names in ocean_meta
OA_clark2 <- OA_clark2[names(ocean_meta)]

# Check columns are in same order
colnames(ocean_meta) == colnames(OA_clark2) # Should return all TRUE

# Bind the two dataframes
final <- rbind(ocean_meta, OA_clark2)
```

### 4. Correctly calculate the log response ratio (lnRR) effect size for every row of the dataframe using metafor’s escalc() function.
```{r}
library(orchaRd)
library(metafor)
rom <- metafor::escalc(measure = "ROM", 
                        m1i = ctrl.mean,
                        m2i = oa.mean,
                        sd1i = ctrl.sd,
                        sd2i = oa.sd,
                        n1i = ctrl.n,
                        n2i = oa.n,
                        data = final,
                        var.names = c("ROM", "ROM_V"))
rom <- mutate(rom, residual = 1:n()) # Added a column for observations / residuals 

plot(rom1$residual, rom1$ROM_V) # Observed outliers in ROM_V, deleted them for further analysis
which(rom$ROM_V > 300) # Set preliminary ylim to 300 to find outliers
rom[c(163, 220, 673),24] # These observations have values: 21969.18, 143043506.14 and 6642701.26
rom1 <- rom[-c(163, 220, 673),] # Made new variable with deleted outliers
```

### 5. Correct meta-analytic model fitted to the data that controls for the sampling variance of lnRR. The model should include a random effect of study and observation. Use metafor’s rma.mv() function.
```{r}
MLMA <- metafor:: rma.mv(ROM ~ ROM_V , V = ROM_V,
                         method = "REML",
                         random=list(~1 | Study,
                                     ~1 | Species,
                                     ~1 | residual),
                         test = "t", dfs = "contain", 
                         data = rom1)
```
(Look at meta-analysis-student.Rmd row 205 for notes)

### 6. Written paragraph of the findings and what they mean which is supported with a figure. The paragraph should include:
  + Correct presentation and interpretation of overall meta-analytic mean and measures of uncertainty around the mean estimate (e.g., 95% confidence intervals).
  + Measures of heterogeneity in effect size estimates across studies (i.e., I2 and/or prediction intervals - see predict() function in metafor)
  + Forest plot showing the mean estimate, 95% confidence interval, and prediction interval with clearly labelled axes, number of samples and studies plotted on figure
```{r eval=F, echo=T}
summary(MLMA) # To get intercept and 95% Confidence intervals
predict(MLMA) # To get 95% Prediction interval
I2_val <-  orchaRd::i2_ml(MLMA) 
I2_val # Shows dissection of variation in effects within and across studies

# Made a new multilevel meta-regression model to observe ROM against moderator "Climate..FishBase."
MLMR <-  metafor:: rma.mv(ROM ~ Climate..FishBase., V = ROM_V,
                          method = "REML",
                          random=list(~1 | Study,
                                      ~1 | Species,
                                      ~1 | residual),
                          test = "t", dfs = "contain", 
                          data = rom1)

orchaRd::orchard_plot(MLMR, mod = "Climate..FishBase.", group = "Study", data = rom, xlab = "Log response ratio effect size (lnRR)", angle = 45) +
  annotate(geom = "text", x = 1, y = -10, 
           label = paste0("italic(I)^{2} == ",round(I2_val[1], 2)), 
           color = "black", parse = TRUE, size = 5) + 
  annotate(geom = "text", x = 0.95, y = -7.8, label = "%", color = "black", parse = FALSE, size = 5) 
```
The intercept was found to be -0.0563. This reflects the overall estimate of the effect from all experiments published in the literature. In other words, it combines the meta-analytic mean effect sizes across all studies. Thus, the intercept infers that there is a weak negative correlation between elevation of CO2 levels and fish behavior across all studies. 

The resultant 95% confidence and prediction intervals are -0.3041 to 0.1916 and -4.2487 to 4.1361, respectively. While 95% of the time we can be confident the true mean would fall between LnRR values of -0.3041 to 0.1916, we also expect effect sizes to range from -4.2487 to 4.1361 with 95% of repeated experiments. Such a broad prediction interval is indicative of high inconsistency between studies. 

Aside from prediction intervals, we can also deduce levels of heterogeneity through I2. I2 represents the allocation of variation from within and across studies into "Total" and the aforementioned random effects. Evidently, we observe complete heterogeneous effect size because sampling variance contributes 0% (100% - I(total)^2) of total variation in effect size. Likewise, we observe 5.60% of total effect size variation due to differences in studies, 4.87% due to differences in species, and 89.53% due to residual effects. 

The MLMR model and Orchard plot relates the effect sizes with the climate, evaluating whether lnRR value changes depending on the climate conditions. Thus, the intercept of MLMR explains the relationship between CO2 levels and fish behavior in correspondence with climate. As indicated on the Orchard plot as well, we observe majority of points with very low precision, with 95% confidence interval intersecting x=0, suggesting no effect. 

### 7. Funnel plot for visually assessing the possibility of publication bias.
```{r}
par(oma = c(3,3,3,3))
metafor::funnel(x = rom1$ROM, vi = rom1$ROM_V, 
                yaxis = "seinv", digits = 2, 
                level = c(0.1, 0.05, 0.01), 
                shade = c("white", "gray55", "gray 75"), las = 1, 
                xlab = "Correlation Coefficient (r)", legend = TRUE)

library(ggplot2)
ggplot(rom, aes(x=tanh(ROM), y = 1/sqrt(ROM_V))) + geom_point(aes(colour = Year..online.), alpha=0.5) + geom_vline(aes(xintercept = 0)) +
    labs(y = "Precision (1/SE)", x = "Correlation Coefficient (r)") + theme_bw()
  # No publication bias exists -> fairly symmetrical and funnel shaped
    # Sampling variance expected to decrease (or the precision increase) when the sample size, and thus power, increases.

yaxis="vi" for the sampling variances,

yaxis="seinv" for the inverse of the standard errors,

yaxis="vinv" for the inverse of the sampling variances,

yaxis="ni" for the sample sizes,

yaxis="ninv" for the inverse of the sample sizes,

yaxis="sqrtni" for the square root of the sample sizes,

yaxis="sqrtninv" for the inverse square root of the sample sizes,

yaxis="lni" for the log of the sample sizes,

yaxis="wi" for the weights.
```

### 8. Time-lag plot assessing how effect sizes may or may not have changed through time
```{r}
ggplot(rom1, aes(y = ROM, x = Year..online., size = 1/sqrt(ROM_V))) + geom_point(alpha = 0.3) +
    geom_smooth(method = lm, col = "red", show.legend = FALSE) + labs(x = "Publication Year",
    y = "Log response ratio (lnRR)", size = "Precision (1/SE)") +
    theme_classic()
  # Small points = Effects with low precision or high sampling varaince
  # smoothing line does not increase much and stays around 0
    # Weak positive relationship between lnRR effect size and year
    # Consistently have low precision and high sampling variance throughout the years
    # Early publications have much lower effect size, while have some outliers (high) around 2014
```

### 9. Formal meta-regression model that includes year as a moderator (fixed effect) to test for time-lag bias
```{r}
MLMR1 <-  metafor:: rma.mv(ROM ~ Year..online., V = ROM_V,
                          method = "REML",
                          random=list(~1 | Study,
                                      ~1 | Species,
                                      ~1 | residual),
                          test = "t", dfs = "contain", 
                          data = rom1)
r2_1 <- orchaRd::r2_ml(MLMR1)
     # R2_marginal R2_conditional 
     #    0.0195         0.1068
    # Time-lag explains 1.95% of the variation in lnRR
```

### 10. Formal meta-regression model that includes inverse sampling var (i.e., 1vlnRR) to test for file-drawer biases
```{r}
rom <- rom %>% mutate("1vlnRR" = 1/(ROM_V))
MLMR2 <-  metafor:: rma.mv(ROM ~ Year..online. + `1vlnRR`, V = ROM_V,
                          method = "REML",
                          random=list(~1 | Study,
                                      ~1 | Species,
                                      ~1 | residual),
                          test = "t", dfs = "contain", 
                          data = rom)
r2_2 <- orchaRd::r2_ml(MLMR2)
rom$`1vlnRR`
```

11. A written paragraph that discusses the potential for publication bias based on the meta-regression results. What type of publication bias, if any, appears to be present in the data? If publication bias is present, what does it mean and what might be contributing to such bias?
12. Identify any studies contributing to publication bias. How do your updated meta-analysis results compare with a meta-analysis by Clement et. al. (2022)? Are there any concerns about these studies? If so, describe using references to existing papers what concerns have been raised?

Reproducibility (30%):
1. Code, analysis and data is setup in a public GitHub Repository
2. Workflow, meta-data and file structure is clearly documented (using a README file) as is the history of changes for each file.
3. Rmarkdown documents follow reproducibility principles:
  + Rmarkdown document rendered as an html!
  + Use Figure and Table code chunks that are referenced in text.
  + Writing of findings is done using inline code chunks with reference to specific object values.
```{r}

```

  
Coding, Writing Structure & Presentation (20%):
1. Code is clearly annotated, clean, and only what is needed is presented
2. Figure and Tables have clear and well labelled captions that are informative and correctly referenced within the document.
3. Sentences are clear and understandable.

Glossary
```{r}
OA_data <- read_csv("./data/Assignment2/OA_activitydat_20190302_BIOL3207.csv")
OA_summary <- OA_data %>% group_by(species, treatment) %>%
    summarise(Mean = mean(activity, na.rm = TRUE),
              SD = sd(activity, na.rm = TRUE),
              N = length(unique(animal_id))) %>%
  rename(Species = "species")

# 2. Merge summary stats of (1) with clark_paper_data.csv
clark_data<- read_csv("./data/Assignment2/clark_paper_data.csv")
OA_clark <- cbind(clark_data, OA_summary)

#3. Merge output from 2 into a larger meta-analysis dataset (i.e., ocean_meta_data.csv)
ocean_meta <- read.csv("./data/Assignment2/ocean_meta_data.csv")
OA_clark1 <- pivot_wider(OA_clark, names_from = treatment,
                     names_glue = "{treatment}_{.value}",
                     values_from = c("Mean", "SD", "N"))

# Reorder col names based on names in meta_data_full
library(dplyr)
OA_clark2 <- OA_clark1 %>% relocate(Species, .after = "Effect type")
names(ocean_meta) <- names(OA_clark2)

# Check columns are in same order
colnames(ocean_meta) == colnames(OA_clark2) # Returns all TRUE

# Bind the two dataframes
final <- rbind(ocean_meta, OA_clark2)
OA_clark1 <-  mutate(OA_clark, ctrl.n = case_when(treatment == "control" ~ N),
                     OA_clark, ctrl.mean = case_when(treatment == "control" ~ Mean),
                     OA_clark, ctrl.sd = case_when(treatment == "control" ~ SD),
                     OA_clark, oa.n = case_when(treatment == "CO2" ~ N),
                     OA_clark, oa.mean = case_when(treatment == "CO2" ~ Mean),
                     OA_clark, oa.sd = case_when(treatment == "CO2" ~ SD))
OA_clark1 <- OA_clark1[,-c(17:20)] # Got rid of columns that do not align with ocean_meta_data

## Used a function to realign the variables for last couple columns
shift <- function(x, n){
  c(x[-(seq(n))], rep(NA, n))}

OA_clark1[,17] <- shift(OA_clark1[,17], 1)
OA_clark1[,18] <- shift(OA_clark1[,18], 1)
OA_clark1[,19] <- shift(OA_clark1[,19], 1)

## Got rid of NAs
OA_clark1 <- OA_clark1[complete.cases(OA_clark1[ , c(17:22)]),]

## Do some renaming of colnames so they match ocean_meta
OA_clark2 <- OA_clark1 %>% rename (CO2_mean = oa.mean,
                                   CO2_sd = oa.sd,
                                   CO2_n = oa.n,
                                   Control_mean = ctrl.mean,
                                   Control_sd = ctrl.sd,
                                   Control_n = ctrl.n)

# Reorder col names based on names in meta_data_full
ocean_meta$Species # = pos11
OA_clark2$species # = pos 16
library(dplyr)
OA_clark2 %>% relocate(species, .after = "Effect type")



```
